---
title: "L3 Raster data"
output: html_document
date: '2024-01-22'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

(Install and) load geodata library for geographical data. The geodata package facilitates access to climate, elevation, soil, crop, species occurrence, and administrative boundary data, and is a successor of the getData() function from the raster package. More info see [https://github.com/rspatial/geodata]

Install package terra and load it. Note geodata depends on terra. Version required: R (>= 3.5.0), terra (>= 1.6.41).

```{r}
#install.packages("geodata")
#install.packages("terra")
library(geodata)
```

use the worldclim_country() function of the geodata package (Hijmans et al. 2023) to download global temperature data from the WorldClim database. The function for WorldClim downloads monthly average climate data. 

```{r}

library(terra)
r <- geodata::worldclim_country(country = "Spain", var = "tavg",
                                res = 100, path = tempdir())

# monthly average temperature in degree Celsius by specifying the country (country = "Spain"), the variable mean temperature (var = "tavg"), the resolution (res = 100), and the path where to download the data to as a temporary file (path = tempdir()). Figure 4.2 shows maps of the monthly average temperature in Spain.
plot(r)
```
```{r}
r
```

average over layers (months)
```{r}
r <- mean(r)
plot(r)
```

remember the "world" sf data we used last lecture, we can get the extent of Spain.
```{r}

library(spData)
library(sf)
names(world)
spain1 = world[world$name_long == "Spain", 1 ]

plot(spain1)
```
get the extent and use crop() to remove the part of the raster that is outside the spatial extent 
```{r}
sextent <- terra::ext(spain1)
r <- terra::crop(r, sextent)
plot(r)
```
We can use the mask() function to convert all values outside the map to NA
```{r}
# Masking
r <- terra::mask(r, vect(spain1))
plot(r)
```

Aggregate groups of cells of a raster  to create a new raster with a lower resolution (i.e., larger cells)

```{r}
coarser <- terra::aggregate(r, fact = 20, fun = "mean", na.rm = TRUE)
plot(coarser)
```

Read your own ".tif" or ".shp" files

```{r}
system.file("ex/elev.tif", package = "terra") # the directory of a file, here is an example using a file that is downloaded once you downloaded the package
```


```{r}
fras =rast(system.file("ex/elev.tif", package = "terra"))
fras
plot(fras)

```
Reading a vector (here shapefile) file.
```{r}
fvec =vect(system.file("ex/lux.shp", package = "terra"))
fvec
plot(fvec)
```

Obtain the centroids of the division polygons, and their coordinates.
```{r}
centroids(fvec)
plot(centroids(fvec))
cp=crds(centroids(fvec))
cp
```
Overlay the raster, shapefile, and centroids. 
```{r}
plot(fras)
plot(fvec, add = TRUE)
points(cp)
```

we can obtain the values of the raster at points using **extract()**. The first argument is a **raster** object, and the second argument a data frame with the points.
```{r}
cpdf <- as.data.frame(cp)
valuesatpoints <- extract(fras, cpdf, xy =TRUE) #xy = True means to add the coordinates in the dataframe. 
head(valuesatpoints) # print the first 5 rows of the table. 
```

Calculate average elevation of each polygon.
```{r}
pavg <- extract(fras, fvec, mean, na.rm = TRUE)$elevation
pavg 
```


More about neighbourhood functions and kernels: A kernel is a small matrix (or array) of numbers that defines a weighted window used in operations like convolution. It is applied to each pixel (or cell) of an image or raster to compute a new value based on the values of neighboring pixels.
```{r}
 

# Create a sample raster
# Define a 3x3 focal window
focal_window <- matrix(1/9, nrow = 3, ncol = 3)

# Apply a focal mean
focal_mean_raster <- focal(fras, w = focal_window, fun = mean)

# View the result
plot(focal_mean_raster)
```
```{r}
# Apply a convolution filter (e.g., Sobel filter for edge detection)
sobel_filter <- matrix(c(-1, 0, 1, -2, 0, 2, -1, 0, 1), nrow = 3, ncol = 3)
neighborhood_raster <- focal(fras, w = sobel_filter, fun = sum)

# View the result
plot(neighborhood_raster)
```
```{r}
gaussian_kernel <- matrix(c(1, 2, 1, 2, 4, 2, 1, 2, 1) / 16, nrow = 3, ncol = 3)

# Apply the kernel as a convolution filter (a neighborhood function)
filtered_raster <- focal(fras, w = gaussian_kernel, fun = sum)
```


```{r}
# View the result
plot(filtered_raster)
```

 
(Advanced) Visualise using ggplot2 and tidyterra:
```{r}
#install.packages("ggplot2")
#install.packages("tidyterra")
library(ggplot2)
library(tidyterra)
fvec$avg =pavg
ggplot(data= fvec) +geom_spatvector(aes(fill = avg)) +scale_fill_terrain_c()
```

More about making maps in R [https://www.paulamoraga.com/book-spatial/making-maps-with-r.html]



# Exercise
1. read the document of **worldclim** climate data download function in geodata [https://rdrr.io/cran/geodata/man/worldclim.html] and then download monthly averaged  total precipitation of Germany, at 1000 m resolution. How large is the dataset? 
2. plot the first layer (the averaged precipitation of January)
3. create a map for annual average total precipitation (average over months and), plot the map. 
4. Download the administrative boundary of Germany with the first level of subdivition. (HINT: You can use **gadm** from **geodata** package: Ger = gadm("Germany", level =1, path = tempdir()))

5. Only show the map to the extent of Germany (that is, only show the precipitation in Germany, outside Germany should be masked out).
6. Show the map that include the annual precipitation, first level administrative boundary of Germany (level at Bundesland/state), and the centroid of each polygon. 

7. Extract the annual averaged precipitation value at the centroid of each Bundesland (state).

8. Calculate the mean annual averaged precipitation over each bundesland. 
 
9. Use an Raster image (you can use the ones used in the lectures or one of your own)
9.1. Perform a smoothing, describe your methods for smoothing and why you chose the method. 
9.2. Implement sobel kernel convolusion in left-right and up-down directions. 



```{r}
library(geodata)
library(terra)
library(ggplot2)
library(tidyterra)

```

```{r}
prec <- geodata::worldclim_country(country = "Germany", var = "prec",
                                res = 1000, path = tempdir())
prec
```
```{r}
plot(prec$DEU_wc2.1_30s_prec_1)
```

```{r}
mprec = mean(prec)
plot(mprec)
Germany = world[world$name_long == "Germany", 1 ]
#cmprec=crop(mprec,Germany)
 
cprec <- terra::mask(mprec, vect(Germany))

Ger = gadm("Germany", level =1, path = tempdir())
cprec2 <- terra::mask(mprec, Ger)
plot(cprec2)
```
```{r}

plot(Ger)
```

```{r}
cen=centroids(Ger)
plot(cprec2)
plot(Ger, add=T)
plot(cen, add=T)
```
```{r}

Ger$pavg <- extract(cprec2, Ger, mean, na.rm = TRUE)$mean
ggplot(data= Ger) +geom_spatvector(aes(fill = pavg)) +scale_fill_terrain_c()
```
